Concurrent Programming in Harmony

### 2020-11-26 book/paper.tex
Latest

      processes with private variables that communicate through message passing}

@@

    instructions by a process is executed indivisibly and cannot be interleaved with 
    machine instructions of another process}

@@

  description={is when a set of processes execute in rounds, waiting for one another

@@

\newglossaryentry{blocked process}

@@

  name=blocked process,
  description={is a process that cannot change the state or terminate or
      can only do so after another process changes the state first.
      For example, a process that is waiting for a lock to become available}

@@

  description={(aka spin-waiting) is when a process waits in a loop for some

@@

  description={(aka parallel execution) is when there are multiple processes executing and

@@

  description={a variable that keeps track of which processes are waiting for a

@@

  description={(aka continuation) describes the state of a running process,

@@

      when a process is allowed to enter the critical section}

@@

  description={(aka critical region) is a set of instructions that only one process

@@

    atomically, as other processes can continue to execute and access shared

@@

  description={is when there are two or more processes waiting indefinitely for one

@@

  description={is when each process eventually can access each resource it needs to access

@@

  reachable state of a process}

@@

  description={an object that can be owned by at most one process at a time.  Useful

@@

  description={is an atomic operation on the Harmony virtual machine, executed by a process}

@@

  description={is the property that two processes never enter the same critical section}

@@

      guaranteed in a bounded number of steps even if other processes are not making progress}

@@

\newglossaryentry{process}

@@

  name=process,

@@

      between processes and threads.  A process has a current context and

@@

\newglossaryentry{process variable}

@@

  name=process variable,
  description={is a variable that is private to a single process and stored in

@@

  description={is a synchronization problem whereby one or more producing processes
    submit items and one or more consuming processes want to receive them.  No item

@@

    properties where ``something good eventually happens,'' like processes being

@@

  description={describes when multiple processes access shared state concurrently,

@@

  description={is a lock on a resource that can be held by multiple processes if they all 

@@

    but blocks the process until the counter is larger than zero first}

@@

  description={is when there is just one process executing, as opposed to concurrent

@@

      shared between multiple processes, as opposed to a process variable}

@@

  description={is an implementation of a lock whereby a process loops until the
    lock is available, at which point the process atomically obtains the lock}

@@

  description={is a model of computing where the state of a process is kept on

@@

  description={is when a process cannot make progress because it is continuously
    losing a competition with other processes to get access to a resource}

@@

  description={is the execution of a machine instruction by a process, updating

@@

        other processes)}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes (aka \emph{threads}\footnote{We
\index{thread}%
do not make a distinction between processes and threads.})
read and update shared variables
\index{shared variable}%
concurrently and synchronize with one another makes programs more
complicated than programs where only one thing happens at a time.

\glsadd{concurrent execution}%
\glsadd{shared variable}%
\glsadd{sequential execution}%
\glsadd{determinism}%
\glsadd{atomicity}%

Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}%
program is mostly \emph{deterministic}.
\index{determinism}%
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
On the other hand,
the output of running concurrent programs depends on how the
execution of the various processes are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}%
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically} (indivisibly)
\index{atomicity}%
because there is no other activity interfering with their execution.
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other processes
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
\index{test}%
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called Harmony that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}%
instead of relying
on luck, Harmony will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger the bug in
the smallest number of steps.

@@

So if model checking does not prove a program correct, why is it
useful?
To answer that question, let us consider a sorting algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set
\{ 1, 2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}%
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}%
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces \emph{Peterson's Algorithm}, an
elegant solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language needed for the rest of the book.
\item \autoref{ch:spinlock} introduces ``hardware'' \emph{locks}
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that supports locks and other synchronization primitives.
\item \autoref{ch:cds} provides an introduction to building concurrent
data structures.
\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
abstraction and implementation of \emph{reader/writer locks},
which allow multiple processes in a critical section if they are
all only executing read-only accesses.
\autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
an easy but undesirable approach to synchronize processes.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:interrupts} discusses how to handle interrupts to a process,
a problem closely related to synchronization among multiple processes.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\end{itemize}

@@

Like Python, Harmony is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in Harmony must be terminated by a semicolon
(even \textbf{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in Harmony.
Correct indentation in Harmony is encouraged but not enforced.
Harmony only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item Harmony does not (currently) support floating point, iterators, or I/O;
Harmony does support \textbf{for} loops and various ``comprehensions.''
\item Python is object-oriented, supporting classes with methods and
inheritance; Harmony has objects but does not support classes.  Harmony supports
pointers to objects and methods.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in Harmony.

@@

\begin{code}
\begin{verbatim}
$ harmony triangle.hny
#states = 13 diameter = 1
#components: 13
no issues found
Open file:///.../harmony.html for more information
$
\end{verbatim}
\end{code}

@@

In Harmony, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states = 103 diameter = 1
#components: 103
no issues found
Open file:///.../harmony.html for more information
$
\end{verbatim}
\end{code}

@@

\glsadd{process}%

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
\autoref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until all \textit{done} flags are set to \texttt{True}.
(\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
After that, it verifies that the value of \textit{count} equals 2.
Like Python, Harmony supports \texttt{assert} statements with two arguments:
the first is a Boolean condition, and the second is a value that is
reported if the condition fails, in this case
the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the processes are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.
And because Harmony tries every possible execution, it will consider
such executions as well.

@@

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}%
in unspecified and often random ways.
A program is correct if it works for any interleaving.
Harmony will try all possible interleavings of the processes
executing machine instructions.

The following is a possible interleaving of incrementers~0 and~1:
\begin{enumerate}
\item incrementer 0 loads the value of \textit{count}, which is 0;
\item incrementer 1 loads the value of \textit{count}, which is still 0;
\item incrementer 1 adds one to the value that it loaded (0), and
stores $1$ into \textit{count};
\item incrementer 0 adds one to the value that it loaded (0), and
stores $1$ into \textit{count};
\item incrementer 0 sets \textit{done}[0] to \texttt{True};
\item incrementer 1 sets \textit{done}[1] to \texttt{True}.
\end{enumerate}

@@

This would fix the issue, but more likely it is the program that must
be fixed.

@@

Again, square brackets and parentheses work the same in Harmony, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[ False, True ]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.  The expression \texttt{[False]} just means
\texttt{False}.

@@

Instead of bits and bytes, a HVM manipulates Harmony values.
A HVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of HVM instructions,
generated from a Harmony program.  The types of instructions will be described later.
\item Shared memory: A HVM has just one memory location containing
a Harmony value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}%
a program counter,
\index{program counter}%
a stack of Harmony values,
and a single mutable general purpose \emph{register}
\index{register}%
that contains a Harmony value.
\end{itemize}

A name tag consists of the name of the main method of the process,
along with an optional tag specified in the \texttt{spawn}
\index{spawn statement}%
\index{tag}%
statement.
The default tag is the first argument to the method.
A name tag is represented as a dictionary with keys \texttt{.name}
and \texttt{.tag}, but we shall often abbreviate it using the notation
\texttt{name/tag}.
For example, in \autoref{fig:inc}, the created processes have name tags
\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the value of
its name tag, program counter, stack, and register.
The state of a HVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_init\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.

\begin{figure}
\begin{code}
\begin{verbatim}
code/Up.hny:1 def incrementer(self):
   0 Jump 25
   1 Frame incrementer self
code/Up.hny:2     count = count + 1;
   2 Load count
   3 Push 1
   4 2-ary +
   5 Store count
code/Up.hny:3     done[self] = True;
   6 Push ?done
   7 LoadVar self
   8 Address
   9 Push True
   10 Store
   11 Return
code/Up.hny:5 def main():
   12 Jump 25
   13 Frame main ()
code/Up.hny:6     await all(done);
   14 Load done
   15 1-ary all
   16 JumpCond False 14
code/Up.hny:7     assert count == 2, count;
   17 AtomicInc
   18 Load count
   19 Push 2
   20 2-ary ==
   21 Load count
   22 Assert2
   23 AtomicDec
   24 Return
code/Up.hny:9 count = 0;
   25 Push 0
   26 Store count
code/Up.hny:10 done = [ False, False ];
   27 Push [False, False]
   28 Store done
code/Up.hny:11 spawn incrementer(0);
   29 Push 0
   30 Dup
   31 Push PC(1)
   32 Spawn
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
\label{fig:inccode}
\end{figure}

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
Harmony values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
\index{directory}%
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}%
of a Harmony value, and it is relative to the
``root'' of the directory tree.
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: \texttt{.count} and \texttt{.done}.  And the value of entry
\texttt{.done} is a dictionary with keys 0 and 1.

\begin{figure}
\begin{code}
\begin{verbatim}
#states = 71 diameter = 5
==== Safety violation ====
__init__/()   [0,25-41] 41 { count: 0, done: [False, False] }
incrementer/0 [1-4]      5 { count: 0, done: [False, False] }
incrementer/1 [1-11]    11 { count: 1, done: [False, True ] }
incrementer/0 [5-11]    11 { count: 1, done: [True,  True ] }
main/()       [13-22]   22 { count: 1, done: [True,  True ] }
>>> Harmony Assertion failed : 1
Open file://.../harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

\begin{figure}
\begin{code}
\harmonysource{csbarebones}
\end{code}
\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two processes.}
\label{fig:csbarebones}
\end{figure}

>>>>

      threads with private variables that communicate through message passing}

@@

    instructions is executed indivisibly by a thread and cannot be interleaved with 
    machine instructions of another thread}

@@

  description={is when a set of threads execute in rounds, waiting for one another

@@

\newglossaryentry{blocked thread}

@@

  name=blocked thread,
  description={is a thread that cannot change the state or terminate or
      can only do so after another thread changes the state first.
      For example, a thread that is waiting for a lock to become available}

@@

  description={(aka spin-waiting) is when a thread waits in a loop for some

@@

  description={(aka parallel execution) is when there are multiple threads executing and

@@

  description={a variable that keeps track of which threads are waiting for a

@@

  description={(aka continuation) describes the state of a running thread,

@@

      when a thread is allowed to enter the critical section}

@@

  description={(aka critical region) is a set of instructions that only one thread

@@

    atomically, as other threads can continue to execute and access shared

@@

  description={is when there are two or more threads waiting indefinitely for one

@@

  description={is when each thread eventually can access each resource it needs to access

@@

  reachable state of a thread}

@@

  description={an object that can be owned by at most one thread at a time.  Useful

@@

  description={is an atomic operation on the Harmony virtual machine, executed by a thread}

@@

  description={is the property that two threads never enter the same critical section}

@@

      guaranteed in a bounded number of steps even if other threads are not making progress}

@@

\newglossaryentry{thread}

@@

  name=thread,

@@

      between threads and threads.  A thread has a current context and

@@

\newglossaryentry{thread variable}

@@

  name=thread variable,
  description={is a variable that is private to a single thread and stored in

@@

  description={is a synchronization problem whereby one or more producing threads
    submit items and one or more consuming threads want to receive them.  No item

@@

    properties where ``something good eventually happens,'' like threads being

@@

  description={describes when multiple threads access shared state concurrently,

@@

  description={is a lock on a resource that can be held by multiple threads if they all 

@@

    but blocks the thread until the counter is larger than zero first}

@@

  description={is when there is just one thread executing, as opposed to concurrent

@@

      shared between multiple threads, as opposed to a thread variable}

@@

  description={is an implementation of a lock whereby a thread loops until the
    lock is available, at which point the thread atomically obtains the lock}

@@

  description={is a model of computing where the state of a thread is kept on

@@

  description={is when a thread cannot make progress because it is continuously
    losing a competition with other threads to get access to a resource}

@@

  description={is the execution of a machine instruction by a thread, updating

@@

        other threads)}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
\index{thread}%
threads read and update shared variables
\index{shared variable}%
concurrently and synchronize with one another makes programs more
complicated than programs where only one thing happens at a time.
%
\glsadd{concurrent execution}%
\glsadd{shared variable}%
\glsadd{sequential execution}%
\glsadd{determinism}%
\glsadd{atomicity}%
%
Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}%
program is mostly \emph{deterministic}.
\index{determinism}%
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
On the other hand,
the output of running concurrent programs depends on how the
execution of the various threads are \emph{interleaved}.
Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}---overhead caused by instrumentation
leads to timing changes that makes such bugs less likely to occur).
\index{Heisenbug}%
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically} (indivisibly)
\index{atomicity}%
because there is no other activity interfering with their execution.
Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other threads
may update memory locations while a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
\index{test}%
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called Harmony that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}%
instead of relying
on luck, Harmony will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Moreover, if the bug is
found, the model checker precisely shows how to trigger the bug in
the smallest number of steps.

@@

So if model checking does not prove a program correct, why is it
useful?
To answer that question, let us consider a sorting algorithm.
Suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set
\{ 1, 2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all \emph{corner cases},
\index{corner case}%
including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.

However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
That said, it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that many correctness properties can be proved using
\emph{invariants}:
\index{invariant}%
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two threads
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces \emph{Peterson's Algorithm}, an
elegant solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language needed for the rest of the book.
\item \autoref{ch:spinlock} introduces ``hardware'' \emph{locks}
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that supports locks and other synchronization primitives.
\item \autoref{ch:cds} provides an introduction to building concurrent
data structures.
\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
abstraction and implementation of \emph{reader/writer locks},
which allow multiple threads in a critical section if they are
all only executing read-only accesses.
\autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
an easy but undesirable approach to synchronize threads.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches threads may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to thread synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of threads are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:interrupts} discusses how to handle interrupts,
a problem closely related to synchronization among multiple threads.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two threads communicate reliably over an unreliable
network.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent threads waiting for one another more than a bounded number of
steps.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\end{itemize}

@@

Like Python, Harmony is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in Harmony must be terminated by a semicolon
(even \textbf{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in Harmony.
Correct indentation in Harmony is encouraged but not enforced.
\item Harmony only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item Harmony does not (currently) support floating point, iterators, or I/O;
Harmony does support \textbf{for} loops and various ``comprehensions.''
\item Python is object-oriented, supporting classes with methods and
inheritance; Harmony has objects but does not support classes.  Harmony supports
pointers to objects and methods.
\end{itemize}
There are also less important ones that you will discover as
you get more familiar with programming in Harmony.

@@

\begin{code}
\begin{verbatim}
$ harmony triangle.hny
#states = 13 diameter = 1
#components: 13
no issues found
$
\end{verbatim}
\end{code}

@@

In Harmony, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states = 103 diameter = 1
#components: 103
no issues found
$
\end{verbatim}
\end{code}

@@

\glsadd{thread}%

Concurrent programming, aka multithreaded programming, involves multiple
threads
\index{thread}%
running in parallel while sharing variables.
\autoref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until all \textit{done} flags are set to \texttt{True}.
(\textbf{await} $c$ is semantically the same as \textbf{while not} $c$: \textbf{pass}.)
After that, method \texttt{incrementer}
verifies that the value of \textit{count} equals 2.
Like Python, Harmony supports \texttt{assert} statements with two arguments:
the first is a Boolean condition, and the second is a value that is
reported if the condition fails, in this case
the actual value of \textit{count}.

The program spawns three threads.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the threads are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.
And because Harmony tries every possible execution, it will consider
such executions as well.

@@

When running multiple threads, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}%
in unspecified and often random ways.
A program is correct if it works for any interleaving of threads.
Harmony will try all possible interleavings of the threads
executing machine instructions.

If the threads run one at a time, then \textit{count} will be incremented
twice and ends up being 2.  However, the following is also
a possible interleaving of incrementers~0 and~1:
\begin{enumerate}
\item incrementer 0 loads the value of \textit{count}, which is 0;
\item incrementer 1 loads the value of \textit{count}, which is still 0;
\item incrementer 1 adds one to the value that it loaded (0), and
stores $1$ into \textit{count};
\item incrementer 0 adds one to the value that it loaded (0), and
stores $1$ into \textit{count};
\item incrementer 0 sets \textit{done}[0] to \texttt{True};
\item incrementer 1 sets \textit{done}[1] to \texttt{True}.
\end{enumerate}

@@

This would fix the issue, but more likely it is the program that must
be fixed, not the specification.

@@

Again, square brackets and parentheses work the same in Harmony, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[False, True]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.  The expression \texttt{[False]} just means
\texttt{False}.

@@

Instead of bits and bytes, a HVM manipulates Harmony values.
A HVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of HVM instructions,
generated from a Harmony program.  The types of instructions will be described later.
\item Shared memory: A HVM has just one memory location containing
a Harmony value.
\item Threads:  Any thread
can spawn an unbounded number of other threads and threads may terminate.
Each thread has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}%
a program counter,
\index{program counter}%
a stack of Harmony values,
and a single mutable general purpose \emph{register}
\index{register}%
that contains a Harmony value.
\end{itemize}

A name tag consists of the name of the main method of the thread,
along with an optional tag specified in the \texttt{spawn}
\index{spawn statement}%
\index{tag}%
statement.
The default tag is the argument to the method.
A name tag is represented as a dictionary with keys \texttt{.name}
and \texttt{.tag}, but we shall often abbreviate it using the notation
\texttt{name/tag}.
For example, in \autoref{fig:inc}, the created threads have name tags
\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.

@@

The state of a thread is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the values of
its name tag, program counter, stack, and register.
The state of a HVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different threads can
have the same context.
The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_init\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.

\begin{figure}
\begin{code}
\begin{verbatim}
code/Up.hny:1 count = 0;
   0 Push 0
   1 Store count
code/Up.hny:2 done = [ False, False ];
   2 Push [False, False]
   3 Store done
code/Up.hny:4 def incrementer(self):
   4 Jump 31
   5 Frame incrementer self
code/Up.hny:5     count = count + 1;
   6 Load count
   7 Push 1
   8 2-ary +
   9 Store count
code/Up.hny:6     done[self] = True;
   10 Push ?done
   11 LoadVar self
   12 Address
   13 Push True
   14 Store
   15 Return
code/Up.hny:8 def main():
   16 Jump 31
   17 Frame main ()
code/Up.hny:9     await all(done);
   18 Load done
   19 1-ary all
   20 JumpCond False 18
code/Up.hny:10     assert count == 2, count;
   21 ReadonlyInc
   22 AtomicInc
   23 Load count
   24 Push 2
   25 2-ary ==
   26 Load count
   27 Assert2
   28 AtomicDec
   29 ReadonlyDec
   30 Return
code/Up.hny:13 spawn incrementer(0);
   31 Push 0
   32 Dup
   33 Push PC(5)
   34 Spawn
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
\label{fig:inccode}
\end{figure}

It may seem strange that there is only one memory location and that each
thread has only one register.  However, this is not a limitation because
Harmony values are unbounded trees.
Both the memory and the register of a thread always contain
a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
\index{directory}%
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}%
of a Harmony value, and it is relative to the
``root'' of the directory tree.
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: \texttt{.count} and \texttt{.done}.  And the value of entry
\texttt{.done} is a dictionary with keys 0 and 1.

\begin{figure}
\begin{code}
\begin{verbatim}
#states = 71 diameter = 5                                                                              ==== Safety violation ====
P0: __init__/()   [0-4,31-43] 43 { .count: 0, .done: [False, False] }
P1: incrementer/0 [5-8]        9 { .count: 0, .done: [False, False] }
P2: incrementer/1 [5-15]      15 { .count: 1, .done: [False, True ] }
P1: incrementer/0 [9-15]      15 { .count: 1, .done: [True,  True ] }
P3: main/()       [17-27]     27 { .count: 1, .done: [True,  True ] }
>>> Harmony Assertion (file=code/Up.hny, line=10) failed: 1
Open file:///Users/rvr/github/harmony.new/harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

\begin{figure}
\begin{code}
\harmonysource{csbarebones}
\end{code}
\caption{\harmonylink{code/csbarebones.hny} A bare bones critical section with two threads.}
\label{fig:csbarebones}
\end{figure}
